---
title: "SBR report"
subtitle: "Activity economic census"
echo: true
number-sections: true
execute:
  warning: false     # Supprime les warnings globalement
  message: false
format: 
# To get prettier table 
  html:
    df-print: paged # Options : paged, kable, tibble, or dt
# To add template
  clean-revealjs: default
author:
  - name: Your Name
    email: alias@email.com
    affiliations: Bureau of Statistics, Lesotho
date: last-modified
---

# Preliminary steps

```{r}
#| output: false
#| code-fold: true
#| code-summary: "URL link"
url <- "https://minio.lab.sspcloud.fr/nrandriamanana/BoS/SBR DATA.csv"
```

```{r}
#| code-fold: true
#| code-summary: "Import packages"
library(readr)
library(dplyr)
library(tidyr)
```

```{r}
#| output: hide
#| code-fold: true
#| code-summary: "Import csv data"
#| message: false
#| warning: false

# 1. Import data

sbr_data <- read_csv(url, locale = locale(encoding = "Windows-1252"), show_col_types = FALSE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
head(sbr_data)
```

```{r}
#| include: false

## **Distribution of establishments by district-2023**

dist_est_district_2023 <- sbr_data %>%
  filter(`Commencement year` == 2023) %>%
  count(DISTRICT_NAME, name = "n_establishments") %>%
  mutate(
    percentage = round(100 * n_establishments / sum(n_establishments), 2)
  ) %>%
  arrange(desc(n_establishments))
```

```{r}
#| include: false

## **Distribution of establishments by district-all years**

dist_est_district <- sbr_data %>%
  count(DISTRICT_NAME, name = "n_establishments") %>%
  mutate(
    percentage = round(100 * n_establishments / sum(n_establishments), 2)
  ) %>%
  arrange(desc(n_establishments))

```

```{r}
#| include: false

## **Distribution of establishments by district-2023**

dist_est_district_2023 <- sbr_data %>%
  filter(`Commencement year` == 2023) %>%
  count(DISTRICT_NAME, name = "n_establishments") %>%
  mutate(
    percentage = round(100 * n_establishments / sum(n_establishments), 2)
  ) %>%
  arrange(desc(n_establishments))
```

```{r}
#| include: false

## **Distribution of establishments by district-2023**

dist_est_district_2023 <- sbr_data %>%
  filter(`Commencement year` == 2023) %>%
  count(DISTRICT_NAME, name = "n_establishments") %>%
  mutate(
    percentage = round(100 * n_establishments / sum(n_establishments), 2)
  ) %>%
  arrange(desc(n_establishments))
```

# Introduction

## Background

The Bureau of Statistics conducted the Census of Establishments as the first phase of the 2024/2025 Economic Activity Census. This phase, carried out from September 2024, aimed to compile a comprehensive list of all formal enterprises operating in various economic sectors for the 2023 financial year. However, activities related to public administration and the informal sector were excluded from the scope of this census.

The census collected key business data, including economic activity classification, employment distribution by gender, turnover, wages, and salaries. The insights derived from this report will serve as a foundation for updated and reliable economic performance indicators aligned with the National Strategic Development Plan II (NSDP II) extension.

# Methodology

# Establishments characteristics

# List of tables

## **Distribution of establishments by district-2023**

```{r}
#| echo: false
show_dist_est_district_2023 <- dist_est_district_2023 %>%
  rename("District name"=DISTRICT_NAME,
         "Total"=n_establishments)

show_dist_est_district_2023
```

## **Distribution of establishments by district- all year**

```{r}
#| echo: false
dist_est_district
```

## **Distribution of establishments by settlement-2023**

```{r}
#| echo: false
#dist_est_settlement_2023
```

## **Distribution of establishments by Employment size-2023**

## **Distribution of licenced and non-licensed establishments by district-2023**

## **Distribution of licensed establishments by settlement-2023**

## **Establishment CEOs/Owner and Employees by Sex-2023**

## **Distribution of establishment by Economic Activity-2023**

## **Distribution of establishments by sector and wage groups-2023**

## **Distribution of establishments by Internet Access-2023**
